---
title: "Annotation of non-DBDs in alternative TF isoforms"
author: "Sviatoslav Sidorov"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    collapsed: no
    fig_align: center
    fig_caption: yes
    highlight: haddock
    keep_md: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r, include=T}
stringsAsFactors = F

library(tidyverse)
library(kableExtra)
```

# Print package versions

Print versions of the loaded libraries:

```{r, include=T}
R.version
packageVersion("base")
packageVersion("tidyverse")
packageVersion("kableExtra")
```

# Read input

Read args:

```{r, include=T}
args = commandArgs()

tf_coding_transcripts_final_ens99_with_fam_names_corrected.with_tsl_name = args[9]
nondbd.annot_name = args[10]
```

# Annotated and summarise non-DBDs

Generate the raw table of non-DBDs:

```{r, include=T}
tf_coding_transcripts_final_ens99_with_fam_names_corrected.with_tsl = readRDS(tf_coding_transcripts_final_ens99_with_fam_names_corrected.with_tsl_name)

tf.matches = tf_coding_transcripts_final_ens99_with_fam_names_corrected.with_tsl

nondbd.annot.raw = tf.matches %>%
  filter(DBD != "Excluded" & DBD != "Unknown") %>%
  filter(is.na(represents_family)) %>%
  filter(!is.na(ipr_accession)) %>%
  dplyr::select(ipr_accession, ipr_description) %>%
  distinct()

write.table(nondbd.annot.raw,
            file = "Other_domains_manually_curated_raw.tsv",
            quote = F,
            sep = "\t",
            row.names = F)
```

Load the table of manually curated non-DBDs. This table was obtained by the manual curation of the table `Other_domains_manually_curated_raw.tsv` generated above:

```{r, include=T}
nondbd.annot = read.delim(nondbd.annot_name)
```

Summarise molecular functions:

```{r, include=T}
nondbd.annot.summary = nondbd.annot %>%
  count(Molecular_function) %>%
  arrange(desc(n))

kable(nondbd.annot.summary) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(width = "100%")  
```

Write down the overall stats table:

```{r, include=T}
write.table(nondbd.annot.summary, 
            file = "Other_domains_manually_curated_summary_mol_function_dbdminus.tsv", 
            quote = F, 
            sep = "\t", 
            row.names = F)
```

Summarise molecular mechanisms:

```{r, include=T}
count_mol_mechanisms = function(mol.mechanism) {
  return(nondbd.annot %>%
           filter(stringr::str_detect(Molecular_mechanism, fixed(mol.mechanism))) %>%
           nrow())
}

mol_mechanism_summary = data.frame(Molecular_mechanism = unique(unlist(stringr::str_split(nondbd.annot %>%
                                                                                            pull(Molecular_mechanism),
                                                                                          fixed("; "))))) %>%
  rowwise() %>%
  mutate(n = count_mol_mechanisms(Molecular_mechanism))

nondbd.type.number = nondbd.annot %>%
  pull(ipr_accession) %>%
  unique() %>%
  length()

mol_mechanism_summary %>%
  mutate(frequency = n / nondbd.type.number) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  mutate(Molecular_mechanism = factor(Molecular_mechanism, levels = Molecular_mechanism)) %>%
  ggplot(aes(x = Molecular_mechanism, y = frequency)) +
    geom_col(fill = "blue") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Summarise all molecular functions of non-DBDs:

```{r, include=T, fig.height=7}
count_mol_functions = function(mol.function) {
  return(nondbd.annot %>%
           filter(stringr::str_detect(Molecular_function, fixed(mol.function))) %>%
           nrow())
}

mol_function_summary = data.frame(Molecular_function = unique(unlist(stringr::str_split(nondbd.annot %>%
                                                                                          pull(Molecular_function),
                                                                                        fixed("; "))))) %>%
  rowwise() %>%
  mutate(n = count_mol_functions(Molecular_function))

nondbd.type.number = nondbd.annot %>%
  pull(ipr_accession) %>%
  unique() %>%
  length()

mol_function_summary %>%
  mutate(frequency = n / nondbd.type.number) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  mutate(Molecular_function = factor(Molecular_function, levels = Molecular_function)) %>%
  ggplot(aes(x = Molecular_function, y = frequency)) +
    geom_col(fill = "red") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

This plot shows the most popular non-DBD types (including non-canonical DBDs).
